Linking Hypotheses and SI-Paradigms

### 2018-02-14 models/linking hypotheses/linkinghypotheses.rmd
reorder the folders

---
title: "Linking Hypotheses and SI_Paradigms"
output: github_document
---

@@

The RSA model of a pragmatically competent speaker is a probability distribution of possible utterances given some world state the speaker intends to communicate to a listener. We'd like to be able to link this model to the categorical responses we observe in the SI_Paradigms experiment. For the purposes of this document, I'll focus on the version of the experiment where speakers rated an utterance on a two-point scale of "Right"/"Wrong".

@@

Consider this output of the model in the context of the following empirical result of the SI_Paradigms pilot: in the condition where a participant is shown a card with two animals X and Y, people overwhelmingly find a guess of "there is an X or a Y on the card" to be "Right"" rather than "Wrong".

@@

Let's first explore a linking hypothesis that makes the wrong predictions in this regard. We might hypothesize that speakers, when confronted with a world state w (i.e. the configuration of animals on the card in a given SI_Paradigms trial) and an utterance u (the guess in that trial), have access to the probability that the S1 distribution assigns to u in w.

@@

The straw-man linking hypothesis described above was clearly inadequate. But it is a linking hypothesis: it generates predictions about categorical behavior (e.g. responses in the SI_Paradigms experiment) from our model; it "maps the response measure onto the theoretical constructs of interest" (Tanenhaus 2000:564-5). The above hypothesis achieves this by positing that the categorical behavior of interest (answering "Right"/"Wrong"") can be understood as a probabilistic process, whereby the probability that a participant responds "Right" when exposed to utterance u in world state w is equal to the speaker proabibility of u in w as computed by the RSA model:

@@

Exploring this requires some transformation to our results. We continue to look at the subset of the SI_paradigms pilot data where response was binary "Right"/"Wrong". Now, for every row in the data (each row corresponds to one response), we add a value S1(u|w) in a new column (speaker_probability) of the data frame, where S1(u|w) is the speaker probability of u (the 'guess' presented on the screen) given w (the configuration of animals on the card).

@@

### Extending the logistic regression model to other SI_Paradigms conditions, e.g. quatenary response

@@

@@

Yet another possibility is that for some utterance u in world state w, P("Right"|u,w) is a function of S1'(u|w), where S1' is a transformation of the pragmatic speaker distribution. More specifically, S1' is a renormalized version of S1 after low-probability utterances are removed from S1 (according to some threshold of 'low' probability).

>>>>

---
title: "Linking Hypotheses and SI-Paradigms"
author: "Brandon Waldon"
output: github_document
---

@@

The RSA model of a pragmatically competent speaker is a probability distribution of possible utterances given some world state the speaker intends to communicate to a listener. We'd like to be able to link this model to the categorical responses we observe in the SI-Paradigms experiment. For the purposes of this document, I'll focus on the version of the experiment where speakers rated an utterance on a two-point scale of "Right"/"Wrong".

@@

Consider this output of the model in the context of the following empirical result of the SI-Paradigms pilot: in the condition where a participant is shown a card with two animals X and Y, people overwhelmingly find a guess of "there is an X or a Y on the card" to be "Right"" rather than "Wrong".

@@

Let's first explore a linking hypothesis that makes the wrong predictions in this regard. We might hypothesize that speakers, when confronted with a world state w (i.e. the configuration of animals on the card in a given SI-Paradigms trial) and an utterance u (the guess in that trial), have access to the probability that the S1 distribution assigns to u in w.

@@

The straw-man linking hypothesis described above was clearly inadequate. But it is a linking hypothesis: it generates predictions about categorical behavior (e.g. responses in the SI-Paradigms experiment) from our model; it "maps the response measure onto the theoretical constructs of interest" (Tanenhaus 2000:564-5). The above hypothesis achieves this by positing that the categorical behavior of interest (answering "Right"/"Wrong"") can be understood as a probabilistic process, whereby the probability that a participant responds "Right" when exposed to utterance u in world state w is equal to the speaker proabibility of u in w as computed by the RSA model:

@@

Exploring this requires some transformation to our results. We continue to look at the subset of the SI-paradigms pilot data where response was binary "Right"/"Wrong". Now, for every row in the data (each row corresponds to one response), we add a value S1(u|w) in a new column (speaker_probability) of the data frame, where S1(u|w) is the speaker probability of u (the 'guess' presented on the screen) given w (the configuration of animals on the card).

@@

### Extending the logistic regression model to other SI-Paradigms conditions, e.g. quatenary response

@@

We can enrich the pragmatic speaker model provided above to model such a function. First, we specify a threshold function which takes as its input a world state w and a probability threshold $\theta$ and returns only those utterances for which S1(u|w) > $\theta$ is true. The following model also specifies a cost function - for now, we'll keep the cost uniform on utterances.

```{r}
model_threshold = "
var states = [['X'], ['Y'], ['Z'], ['X','Y'], ['X','Z'], ['Y','Z'], ['X','Y','Z']];
var utterances = ['X', 'Y', 'XorY', 'XandY', 'Z','XorZ','XandZ','YorZ','YandZ','XandYandZ','XorYorZ'];

var cost = function(u) {
  return 1
};

// set speaker optimality
var alpha = 1

// pragmatic speaker
var speaker = cache(function(state) {
  return Infer({model: function(){
    var utt = utterancePrior()
    factor(alpha * literalListener(utt).score(state))
    return utt
  }})
});

// threshold function

var threshold = function(state,threshold) {
  var supported_utts_string = speaker(state).support().toString();
  var supported_utts = supported_utts_string.split(',');
  return filter(function(x) { return Math.exp(speaker(state).score(x)) > threshold; }, supported_utts)
}
"
```

Now it is possible to model the linking function of interest: we say that for any utterance u given world state w and probability threshold $\theta$, u is "Right" so long as S(u|w) exceeds $\theta$ and is "Wrong" otherwise (more precisely, the model returns "Right" so long as u is in the set of utterances returned by threshold(w,$\theta$), and "Wrong" otherwise).

```{r}
right_or_wrong <- "
var right_or_wrong = function(utt,state,theta) {
  if (threshold(state,theta).includes(utt)) { return 'Right'; } else { return 'Wrong'; }
}
"
```

Suppose the utterance of interest is 'X or Y' in the world state ['X','Y'] given a probability threshold of 0.2. Which categorical response does our model predict?

```{r}
commands <- "right_or_wrong('XorY',['X','Y'],0.2)"
code <- paste(model_threshold,right_or_wrong,commands,sep = "\n")
webppl(code)
```

For the experimental conditions in which more than two choices were provided, we need to specify an enriched linking function. For the ternary choice condition, we need a function that takes as its input an utterance, a world state, and two thresholds $\theta_1$ and $\theta_2$. An utterance is "Wrong" in world state w just in case S1(u|w) is less than $\theta_1$. It is "Right" just in case S1(u|w) > $\theta_2$. It is "Neither" just in case  $\theta_1$ < S1(u|w) < $\theta_2$.

What if our $\theta_1$ is 0.1, and our $\theta_2$ is 0.2? What does this linking model predict a respondent will say about 'X or Y' in world state ['X','Y']?

```{r}
right_neither_wrong <- "
var right_neither_wrong = function(utt,state,thetaone,thetatwo) {
if (threshold(state,thetatwo).includes(utt)) { return 'Right'; }
if (threshold(state,thetaone).includes(utt)) { return 'Neither'; } 
else { return 'Wrong'; }
}
"

commands <- "right_neither_wrong('XorY',['X','Y'],0.1,0.2)"
code <- paste(model_threshold,right_neither_wrong,commands,sep = "\n")
webppl(code)
```

@@

Yet another possibility is that for some utterance u in world state w, P("Right"|u,w) is a function of S1'(u|w), where S1' is a transformation of the pragmatic speaker distribution. More specifically, S1' is a renormalized version of S1 after low-probability utterances are filtered from S1 (according to some threshold of probability).

In order to explore this hypothesis, we first need to model S1' (specified below as speaker_filtered). Speaker_filtered is a function of utterancePrior_filtered - an utterance prior function which takes as its input only high-probability utterances; that is, utterances for which S1(u|w) > $\theta$ is true.

filter_utterances <- "
var utterancePrior_filtered = function(state,theta) {
  return uniformDraw(threshold(state,theta).toString().split(','));
};

var speaker_filtered = cache(function(state,theta) {
  return Infer({model: function(){
    var utt = utterancePrior_filtered(state,theta)
    factor(alpha * literalListener(utt).score(state))
    return utt
  }})
});
"

model_threshold_2 <- paste(model_threshold,filter_utterances,sep = "\n")

Let's explore this new S1' distribution over utterances, given a world state ['X','Y'] and that we are only interested in those utterances for which it is true that S1(u|w) > 0.12.

```{r}
commands <- "speaker_filtered(['X','Y'],0.12)"
code <- paste(model_threshold_2, commands, sep = "\n")
webppl(code)
```

With this model, we could revisit the logistic regression linking function explored above: the log odds of categorical response is a function of the (filtered) speaker probability of that utterance in a particular world state w and given a probability threshold $\theta$.
